Inference of proto-neutron star properties from gravitational wave data in core-collapse supernovae.

### 2020-10-09 introduction.tex
update with final results

The life of massive stars {(those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$)} ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. {Nearby} core-collapse supernova (CCSN) explosions {are expected to be sources of gravitational waves (GW) and are one of the main candidates for the next great discovery by current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slow rotating cores, which are present in the bulk of CCSN progenitors. This event could be detected with advanced ground-based GW detectors within $5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}. For the case of fast rotating progenitor cores the result is likely a magneto-rotational explosion, with } a more powerful signal {that could} be detected {within $50$~kpc and for some extreme models 
up to $5-30$ Mpc \cite{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated to long GRBs 
\citep{Chapman:2007}), making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$. Therefore, we focus this work only in neutrino-driven CCSNe. 
Despite the low rates, CCSN are of great scientific interest because they produce} a complex GW signals which could provide significant clues about the physical processes that occur in the moments after the collapse.

In the {last decade, a significant progress has} been made in the development of numerical codes, {in particular in the treatment of multidimensioal effects \citep{BMueller:2020}.}
% {[PCD: I don't think the next sentence is necessary since we have already discarded magneto-rotational events for this work.]
%\sout{The waveforms produced by the magneto-rotational mechanism in particular is well understood. The core-bounce signal can be directly related with the rotational properties of the core} \cite{Dimmelmeier:2007, abdikamalov:14, Richers:2017}. \sout{However, the low rate of this kind of events and its expected low amplitude in the slow-rotation case, will probably impede its detection.}}
In the case of  neutrino-driven {explosions}, the GW emission is {primarly induced by instabilities developed at the newly formed proto-neutron star (PNS) and by the non-spherical accreting flow of hot matter over its surface \citep{Kotake:2017}.  This} dynamics excite the different modes of oscillation of the PNS{, which ultimately leads to the emission of GWs. The frequency and time evolution of these modes carry information about the
properties of the GW emitter and could allow to perform PNS asteroseismology. }

@@

{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow feature in the post bounce evolution with raising frequency
from about $100$~Hz up to a few kHz (at most) and a typical duration of $0.5-1$~s. This feature has been interpreted as a continuously excited gravity mode (g-mode, see \citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS \cite{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015, Kuroda:2016, Andresen:2017}. In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS.} The {typical} frequencies of {these} modes make them {a} promising {source} for ground-based interferometers {(aLIGO, aVirgo, KAGRA)}.

@@

 {We base this work in the PNS mode analysis performed by \cite{Torres:2018, Torres:2019a}, which explored the eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
 using results from 2D CCSN numerical simulations as a background. Their results show a good match of the mode frequencies computed and the features observed in the GW spectrum of the same simulation (specially
 when space-time perturbations are included \citep{Torres:2019a}).  This result} reveals that it is posible to perform CCSN asteroseismology {under realistic conditions} and serves as a starting point to carry out inference of astrophysical parameters of PNSs.  {\cite{Torres:2019b} went one step further showing that it was possible to derive simple relations} between the { instantaneous frequency of the g-mode and} the mass and radius of the PNS {at each time of the evolution}. These relations are universal in the sense that they do not depend {on the equation of state (EOS) used} or the mass of the progenitor, {and only weakly on} the numerical code used {(see discussion in section~\ref{sec:simulations})}. {Similar relations have been found by \cite{Sotani:2020a,Sotani:2020b}, which also found that the universal relations do not depend on the dimensionality (1D, 2D or 3D) of the numerical simulation used as a background.}

>>>>

The life of massive stars {(those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$)} ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. {Nearby} core-collapse supernova (CCSN) explosions {are expected to be sources of gravitational waves (GWs) and are one of the main candidates for the next great discovery by current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slow rotating cores, which are present in the bulk of CCSN progenitors.
\mab{The emitted GWs} could be detected with advanced ground-based GW detectors \mab{(
  Advanced LIGO\citep{TheLIGOScientific:2014jea},
  Advanced Virgo\citep{TheVirgo:2014hva} and
  KAGRA\citep{Aso:2013eba})} within $5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of fast rotating progenitor cores the result is likely a magneto-rotational explosion, with } a more powerful \mab{GW} signal {that could} be detected {within $50$~kpc and for some extreme models 
up to $5-30$ Mpc \cite{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated to long GRBs 
\citep{Chapman:2007}), making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{add ref?}. Therefore, we focus this work only in neutrino-driven CCSNe. 
Despite the low rates, CCSN are of great scientific interest because they produce} a complex GW signals which could provide significant clues about the physical processes that occur in the moments after the collapse.

In the last decade, significant progress has been made in the development of numerical codes, {in particular in the treatment of multidimensioal effects \citep{BMueller:2020}.}
% {[PCD: I don't think the next sentence is necessary since we have already discarded magneto-rotational events for this work.]
%\sout{The waveforms produced by the magneto-rotational mechanism in particular is well understood. The core-bounce signal can be directly related with the rotational properties of the core} \cite{Dimmelmeier:2007, abdikamalov:14, Richers:2017}. \sout{However, the low rate of this kind of events and its expected low amplitude in the slow-rotation case, will probably impede its detection.}}
In the case of  neutrino-driven explosions, the GW emission is {primarly induced by instabilities developed at the newly formed proto-neutron star (PNS) and by the non-spherical accreting flow of hot matter over its surface \citep{Kotake:2017}.  This} dynamics excite the different modes of oscillation of the PNS{, which ultimately leads to the emission of GWs. The frequency and time evolution of these modes carry information about the
properties of the GW emitter and could allow to perform PNS asteroseismology. }

@@

{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow {\mab oscillation} in the post bounce evolution with raising frequency
from about $100$~Hz up to a few kHz (at most) and a typical duration of $0.5-1$~s. This feature has been interpreted as a continuously excited gravity mode (g-mode, see \citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS \cite{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015, Kuroda:2016, Andresen:2017}. In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS.} The {typical} frequencies of {these} modes make them {a} promising {source} for ground-based interferometers.

@@

 % {We base this work in the PNS mode analysis performed by \cite{Torres:2018, Torres:2019a}, which explored
\mab{The eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
using results from 2D CCSN numerical simulations as a background studied in \cite{Torres:2018, Torres:2019a}
show a good match of the mode frequencies computed and the features observed in the GW spectrum of the same
simulation (specially when space-time perturbations are included \citep{Torres:2019a}).}
This result reveals that it is posible to perform CCSN asteroseismology {under realistic conditions} and serves as a starting point to carry out inference of astrophysical parameters of PNSs.  {\cite{Torres:2019b} went one step further showing that it was possible to derive simple relations} between the { instantaneous frequency of the g-mode and} the mass and radius of the PNS {at each time of the evolution}. These relations are universal in the sense that they do not depend {on the equation of state (EOS) used} or the mass of the progenitor, {and only weakly on} the numerical code used {(see discussion in section~\ref{sec:simulations})}. {Similar relations have been found by \cite{Sotani:2020a,Sotani:2020b}, which also found that the universal relations do not depend on the dimensionality (1D, 2D or 3D) of the numerical simulation used as a background.}

### 2020-10-09 results.tex
update with final results

To estimate how accurately we can infer the time evolution of $r=M_{\rm PNS}/R_{\rm PNS}^2$ in the
gravitational wave detector data, we have added \pcd{the} GW signal \pcd{from {\texttt s20S}} to 
100 Gaussian noise realisations whose power spectral density follows \pcd{the} advanced LIGO (aLIGO)
spectrum~\cite{aLIGOsens:2018} shown on Figure~\ref{fig:spectrum}. 
We have varied the distance to the source, covering a large
range of distances for which a detection in second generation of gravitational wave detectors
is feasible. The source is optimally oriented with
respect to the gravitational wave detector. We are assuming a GW signal from a core collapse
phenomena has been identified in the data and that the beginning of the GW signal is known within $O(10~ms)$.
The data (signal embedded in noise) are whitened using the function {\it prewhiten} of the R-package TSA.
An auto-regressive model with maximal \textcolor{red}{100} coefficients has been used.

@@

$\Delta$ values of each of the 100 noise realisation are shown as well as function of the distance
on Figure \ref{fig:s20results}. For a source located up to $\sim$\unit[10]{kpc} the relative error
remains smaller than 20\% \pcd{[PCD: should we comment on the outliers?]}. At small distance
$\Delta$ is small but not null. This reflects the approximation of the model used for $r$.
It is nevertheless remarkable that one can reconstruct the ratio time series with a good
precision at distance up to $\sim$ 9 kpc for this particular waveform, with $coverage$ value
larger than 80\%.
We have tested that the method does not depend on features of {\texttt s20S} using \pcd{the} 7
other waveforms \pcd{of the {\it test set}} described in section \ref{sec:simulations} covering
a large range of progenitor masses.
Figure \ref{fig:aLIGO_cov_allwvf} shows that apart {\tt s11.2--LS220}, the ratio is well
reconstructed for all waveforms up to $\sim$ 9kpc.
To measure the level of contamination of $coverage$ by reconstructed ratios not associated to a
signal, we have run 100 simulations without injecting a signal. The median of $coverage$ as well as
the 95 quartile are shown on Figure \ref{fig:aLIGO_cov_allwvf}. The noise only median value is
significantly different from zero because the g-mode reconstruction algorithm is looking for a
continuously frequency increasing track in the spectrogram, starting between 0 and 200 Hz.
Outliers can also reach values as high as 80\%.
Figure \ref{fig:aLIGO_prec_allwvf} shows $\Delta$ as function of the distance for the same signals
as well as the result when only noise is considered.
In Table \ref{tab:results} we are reporting the distance at which $coverage$ median is lower than
95\% of the noise only values. These numbers are an estimate of the order of magnitude of the 
source maximal distance at which a reconstruction of the ratio could be possible with current
and future generations of gravitational wave detectors. We have checked that $coverage$ and $\Delta$
provide similar values.

\mab{add comment about ADV}

@@

 \caption{Boxplots of $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realisations is considered for each distance.
 \pcd{[PCD: I guess you use "coverage probability" because we are showing the box plots. Shouln't we say $\Delta$ probability as well? Or just "coverage"? Also, 
 how are outliers computed? Why are there so many outliers in $|delta$? If they are a significant fraction of the data points why are they considered outliers?]}}
  \label{fig:s20results}
\end{figure}

\begin{figure}
 \centering
 \includegraphics[width=0.5\textwidth]{plots/spectrum}
 \caption{} \label{fig:spectrum}
\end{figure}

@@

The same analysis has been performed using expected sensitivity curves for the third generation of
gravitational wave detectors. The US based project Cosmic Explorer~\cite{reitze2019cosmic} is proposing
to reach out its design sensitivity circa 2040 through two phases labeled CE1 and CE2 also shown in
Figure~\ref{fig:spectrum}. In Europe, the Einstein Telescope is composed of three nested interferometers
in a configuration of an equilateral triange~\cite{Hild_2011}.

\mab{explain ET config}
\mab{comment figure 9}

@@

\hline
\multirow{3}{*}{aLIGO} & $d_{r}$ & 7   & 27  & 24  & 22  & 16  & 10  & 37  & 45 \\
\cline{2-10}
                       & $d_{det}$ & 11 & 36 & 26 & 27 & 21 & 16 & 74 & 61\\
%                       & SNR      & 14 & 46 & 33 & 35 & 27 & 21 & 96 & 80\\

\hline
\hline
\multirow{3}{*}{ADV}   & $d_{r}$ & 6   & 25  & 20  & 19  & 14  & 10  & 43  & 42 \\
\cline{2-10}
                       & $d_{det}$ &  &  &  &  &  &  &  & \\
%                       & SNR      &  &  &  &  &  &  &  & \\

\hline
\hline
\multirow{2}{*}{CE1}   & $d_{r}$   & 79  & 303 & 257 & 229 & 186 & 115 & 523 & 490 \\
\cline{2-10}
                       & $d_{det}$ & 115 & 377 & 270 & 282 & 217 & 168 & 774  & 633\\
%                       & SNR      & 149 & 490 & 352 & 366 & 282 & 218 & 1006 & 822\\

\hline
\multirow{2}{*}{CE2}   & $d_{r}$  & 135 & 518 & 464 & 405 & 305 & 183 & 391 & 700 \\
\cline{2-10}
                       & $d_{det}$ & 197 & 649 & 468 & 489 & 375 & 294 & 1347  & 1100\\
%                       & SNR    & 256 & 843 & 608 & 635 & 487 & 382 & 1751 & 1430\\

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_all3G}
  \caption{Median of $coverage$ for s20.0--SFHo CCSN waveform embedded in 3G detectors noise and located at different distance from the Earth. } \label{fig:s20--SFHo_all3G}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{} \label{fig:distances}
\end{figure}

>>>>

To estimate how accurately we can infer the time evolution of $r=M_{\rm PNS}/R_{\rm PNS}^2$ in the
gravitational wave detector data, we have added \pcd{the} GW signal \pcd{from {\texttt s20S}} to 
100 Gaussian noise realisations whose power spectral density follows \pcd{the} advanced LIGO (aLIGO)
spectrum~\cite{aLIGOsens:2018} shown on Figure~\ref{fig:spectrum}.

\begin{figure}
 \centering
 \includegraphics[width=0.5\textwidth]{plots/spectrum}
 \caption{Amplitude spectral density of the GW detectors advanced LIGO (aLIGO) and Advanced Virgo (ADV) at design sensitivity and of the proposed third generation detectors Cosmic Explorer and Einstein Telescope. Eintein Telescope sensitivity curve ET-B is obtained pushing second generation detector technology at its limit. ET-C and ET-D sensitivity curves correspond to a detector configuration where a low-power cryogenic low-frequency interferometer and a high-power room temperature high-frequency interferometer are sharing the same infrastructure~\cite{Hild_20211}. Cosmic Explorer design sensitivity will be achieved in two stages. Stage 1 (CE1) is expected to use the technology developed for the “A+” upgrade to advanced LIGO but scaled up to a 40 km detector while stage 2 (CE2) will implement state-of-the-art technology to decrease quantum and thermal noises~\cite{reitze2019cosmic}. } \label{fig:spectrum}
\end{figure}

We have varied the distance to the source, covering a large
range of distances for which a detection in second generation of gravitational wave detectors
is feasible. The source is optimally oriented with
respect to the gravitational wave detector. We are assuming a GW signal from a core collapse
phenomena has been identified in the data and that the beginning of the GW signal is known within $O(10~ms)$.
The data (signal embedded in noise) are whitened using the function {\it prewhiten} of the R-package TSA.
An auto-regressive model with maximal \textcolor{red}{100} coefficients has been used.

@@

$\Delta$ values of each of the 100 noise realisation are shown as well as function of the distance
on Figure \ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%.
At small distance, $\Delta$ is small but not null. This reflects the approximation of the model used for $r$.
It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision at distance up to $\sim$ 9 kpc for this particular waveform, with $coverage$ value
larger than 80\%. There are few noise realisations for a source located at $<$ \unit[9]{kpc} for which
$\Delta$ takes large values, indicating that the method start failing to reconstruct with accuracy the ratio.

@@

 \caption{Boxplots of $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realisations is considered for each distance.}
  \label{fig:s20results}
\end{figure}

We have tested that the method does not depend on features of {\texttt s20S} using the 7
other waveforms of the {\it test set} described in section \ref{sec:simulations} covering
a large range of progenitor masses.

Figure \ref{fig:aLIGO_cov_allwvf} shows that apart {\tt s11.2--LS220} and to a lesser extent
{\tt s20.0--SFHo}, the ratio is well reconstructed for all waveforms up to $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio
we have run 100 simulations without injecting a signal and have measured $coverage$ for the
reconstructed ratios.
The median of $coverage$ as well as the 95 quartile are shown on Figure \ref{fig:aLIGO_cov_allwvf}.
The noise only median value is null in this case, but it can be different from zero because
the g-mode reconstruction algorithm is looking for a continuously frequency increasing track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effects explains why outliers can reach
values as high as 80\%.
Figure \ref{fig:aLIGO_prec_allwvf} shows $\Delta$ as function of the distance for the same signals
as well as the result when only noise is considered. 
In Table \ref{tab:results} we are reporting the distance $d_r$ at which $coverage$ median is lower
than 95\% of the noise only values. We have checked that $coverage$ and $\Delta$ provide similar
values.
These numbers are an estimate of the order of magnitude of the source maximal distance at which a
reconstruction of the ratio could be possible with current GW detectors.
They are also upper limits as we are taking into account the detector antenna response in our
simulation but consider the source is optimally oriented. 
Table \ref{tab:results} reports also $d_r$ for the advanced Virgo detector at design sensitivity.
Results are very similar to aLIGO, despite the detector sensitivity differences.
Note that Table \ref{tab:results} provides the distance at which one could detect a source
optimally oriented with a matched filter signal-to-noise ratio of 13.

@@

The same analysis has been performed using expected sensitivity curves for the third generation of
gravitational wave detectors. In Europe the Einstein Telescope project proposes to host in a 10-km
equilateral triangle configuration 3 low-power low-frequency cryogenic interferometers as well as 3
high-power high-frequency interferometers. Three sensitivity curves, ET-B, ET-C and ET-D corresponding
to different options and stages of the project~\cite{Hild_2011} are considered in this study.
The US based project Cosmic Explorer~\cite{reitze2019cosmic} is proposing to reach its design
sensitivity circa 2040 through two phases labeled CE1 and CE2 also shown in Figure~\ref{fig:spectrum}.

Figure \ref{fig:s20--SFHo_all3G} shows $\Delta$ as function of the source distance for {\tt s20.0--SFHo}
waveform for the 5 3G detectors configurations. Overall, the ratio is well reconstructed up to distances
in the range \unit{100--200}[kpc] which represents an improved of at least a factor 10 with respect to
advanced LIGO and advanced Virgo detectors. We can also note that the Einstein Telescope results lay in
between the 2 Cosmic Explorer results. This is confirmed for all other waveforms, expect {\tt s25.0-LS220}
for which the maximal distance reach in CE2 is significantly lower than CE1. This is partly due to the
small variation of the reconstruction quality to the distance of the source making the estimation of $d_r$
rather incertain for this waveform. All results are summarized in Table \ref{tab:results} and Figure
\ref{fig:distances}. It is remarkable that with 3G detectors the ratio could be reconstructed for sources
located up to several hundred of kpc. It is nevertheless important to note the rather large range obtained
for the different waveforms probing a large range of progenitor masses.

\mab{add comment on the mass progenitor dependence}\\
\mab{say that the main depend is the SNR as shown by ddet}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_all3G}
  \caption{Median of $\Delta$ for s20.0--SFHo CCSN waveform embedded in 3G detectors noise and located at different distance from the Earth. } \label{fig:s20--SFHo_all3G}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{Maximal distance $d_{r}$ at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for a source optimally oriented with respect to the GW detectors for the 7 CCSN waveforms considered in this study.} \label{fig:distances}
\end{figure}

@@

\hline
\multirow{3}{*}{aLIGO} & $d_{r}$   & 7 & 28 & 24  & 22 & 16 & 11 & 38 & 46 \\
\cline{2-10}
                       & $d_{det}$ & 11 & 36 & 26 & 27 & 21 & 16 & 74 & 61\\
%                      & SNR      & 14 & 46 & 33 & 35 & 27 & 21 & 96 & 80\\

\hline
\hline
\multirow{3}{*}{ADV}   & $d_{r}$   & 7  & 26  & 20 & 19 & 15 & 10 & 43 & 42 \\
\cline{2-10}
                       & $d_{det}$ &  10 & 32 & 22 & 23 & 18 & 13 & 64 & 52\\
%                      & SNR      &  13 & 41 & 29 & 30 & 23 & 17 &  83 & 68\\

\hline
\hline
\multirow{2}{*}{CE1}   & $d_{r}$   & 79  & 304 & 258 & 229 & 187 & 115 & 524 & 490 \\
\cline{2-10}
                       & $d_{det}$ & 115 & 377 & 270 & 282 & 217 & 168 & 774  & 633\\
%                      & SNR      & 149 & 490 & 352 & 366 & 282 & 218 & 1006 & 822\\

\hline
\multirow{2}{*}{CE2}   & $d_{r}$  & 135 & 499 & 451 & 405 & 305 & 183 & 391 & 898 \\
\cline{2-10}
                       & $d_{det}$ & 197 & 649 & 468 & 489 & 375 & 294 & 1347  & 1100\\
%                      & SNR    & 256 & 843 & 608 & 635 & 487 & 382 & 1751 & 1430\\

@@
